A novel route to the spontaneous formation of porous crystals via viscoelastic phase separation

### 2016-03-12 crystal-gel/nature/crystal_gel2.tex
Update sigma, R and volume fractions. More honest description of how we get them.

In our experimental setup, the sample cell is put in contact with a reservoir of a salt solution through a semi-permeable membrane (see Methods 
on the details of our experiments).
At $t=0$, the increase in the concentration of salt ions within a few Brownian times ($\tau_B$) screens the Coulombic repulsion between colloidal particles,
which are then subject to attractive depletion forces, making the system thermodynamically unstable and leading to phase separation. 
%In the following we discuss all the stages of the formation of the crystal gels: phase separation, stress-driven agein and crystallization.
The experimental data are taken at two different volume fractions ($\phi\approx 0.13$ and $\phi\approx 0.30$) and for different values
of the polymer concentration, $c_p=0.38,0.48,0.57,1.07$ mg/g for $\phi\approx 0.30$ and $c_p=0.82,1.36$ mg/g for $\phi\approx 0.13$.
In Fig.~\ref{fig:phase_separation}a, we superimpose the state points with the phase diagram, which we calculated from generalised free volume theory~\cite{Fleer2008} using the experimental values for the size of the colloids, the radius of gyration of the polymer, and the polymer concentration (see Methods).
For each state point, scans were made at early times (every $10$~s, i.e. $4\tau_B$) and at the late stage (every $30$~s) of the gelation process.

@@

As described above and also in Methods, salt injection initiates liquid-gas phase separation of the colloidal suspension. 
All samples share the same early stages of spinodal decomposition. 
A typical phase ordering process observed at $\phi\approx 0.30$ and $c_p=0.38$ mg/g can be seen in Supplementary Movies 1 and 2. 
Due to strong dynamical asymmetry between colloids and the solvent \cite{tanaka1999colloid}, the colloidal particles start aggregating, 
eventually forming a percolating network. Thus, a dense network (liquid) coexists with freely diffusing monomers (gas).
In Fig.~\ref{fig:phase_separation}a, the state points which did not phase separate are indicated with open triangles, while
the other symbols indicate samples undergoing phase separation. The (black) continuous line is the gas spinodal computed from free volume theory,
and it is in good agreement with our experimental points (the small discrepancy at small colloidal volume fractions 
is common in free volume theory)~\cite{Royall2007,lu2008gelation}. 
The topology of the network is similar to all our samples at the initial stages. 
This is shown for example in Fig.~\ref{fig:phase_separation}b, where we plot the radius of gyration of colloidal clusters ($R_g$)
as a function of the cluster size ($s$) at the percolation point. Two particles belong to the same cluster if their distance is
within the first peak of the radial distribution function. All samples show the same scaling law, $R^2_{\rm g}\sim s^{2/D}$, which is compatible with
the random gelation universality class exponent ($D=2.53$) as shown by the dashed line in Fig.~\ref{fig:phase_separation}b
(we do not imply that the formation of the network follows the random gelation universality class, just that our
fractal dimension is compatible with it).

@@

While sharing similar structural properties at early times, the samples show significantly different behaviours in the later stages of the phase separation process.
To examine this in more detail, we plot in Fig.~\ref{fig:phase_separation}c the structure factor for the final stages of the gelation process for all state points.
Calculation of the structure factor is done with the Hanning window function, to minimize boundary effects.
The figure shows that at low wavenumber $q$, the structure factor displays fractal scaling compatible with Guinier law, $S(q)\sim q^{-D}$.
But a difference in the fractal dimension $D$ between the $\phi\approx 0.30$ and $c_p=0.38$ mg/g state point and other state points starts to become visible. In fact, while states
with high polymer concentration retain the exponent $D=2.5(3)$, 
which is the random gel universality class exponent, as we also observed in the early stages of the gelation process (see Fig.~\ref{fig:phase_separation}b),
the state point with low polymer concentration ($\phi\approx 0.30$ and $c_p=0.38$ mg/g) displays the biggest deviation from that exponent. In the figure we also
plot the exponent $D=3$, which corresponds to volume growth, which we will show in the following analysis to be the correct exponent for this state. 
The difference is due to the onset of crystallization in the low-polymer concentration sample. This is already evident in the high $q$ behaviour of
the structure factors (Fig.~\ref{fig:phase_separation}c), where the diffraction planes appear as sharper peaks for $\phi\approx 0.30$ and $c_p=0.38$ mg/g.

To gain additional insight, in Fig.~\ref{fig:phase_separation}d we plot the average local volume fraction as a function of the number of neighbours.
\textcolor{red}{The volume fraction is obtained by computing the Voronoi diagram of each configuration,
which uniquely assigns a volume to each colloidal particle.}
For all state points, we can see that the average volume fraction of a colloidal
particle increases with its number of neighbours. Approaching close packing (12 neighbours), we clearly see two families of curves. The state point
at $\phi\approx 0.30$ and $c_p=0.38$ mg/g reaches close packing at a volume fraction of around $\sim 68\%$, which is in reasonable agreement with the
volume fraction of the stable crystal phase~\cite{ilett1995phase}. All other state points, instead, reach close packing at a volume fraction $\sim 64\%$, which is
indeed close to the volume fraction of the attractive glass state~\cite{pham2002multiple}. 
%(\textbf{the absolute values of these volume fraction are not totally convincing, probably because of a not optimal estimate of the size of the colloids. We should run the code of Mathieau with size detection}). 
This result clearly indicates that the state point at $\phi\approx 0.30$ and $c_p=0.38$ mg/g follows a different arrest mechanism, in which phase separation
is arrested by crystallization and not by glassiness. Although our estimation of the absolute volume fraction may involves systematic errors of a few \%, the trend should be robust~\cite{Poon2012}.

@@

A direct analysis of the colloidal positions at long times confirms that indeed the new structure obtained at
$\phi\approx 0.30$ and $c_p=0.38$ mg/g has different morphological properties than usual colloidal gels.
We directly compare the two structures in Fig.~\ref{fig:network}, where Gaussian filtering is used to depict the network structure as a continuous field.
Panel a presents the results for high
polymer concentration gel, $c_p=1.07$ mg/g; panel b presents the results for the low polymer concentration sample, $c_p=0.38$ mg/g.
The comparison shows that the new arrest mechanism produces a network structure with bigger pores, smoother surfaces, and thicker strands (panel b) compared
to the colloidal gel sample (panel a). The bottom row shows the individual particle positions for slabs with a thickness of 5 particle diameters.
From the figure it is immediately clear that the network strands at low polymer concentration are crystalline. The new arrest mechanism thus
involve the formation of a crystal-gel network. In order to explain the morphological differences between colloidal gels and crystal-gels,
we now consider the kinetics of formation in our samples.

To confirm the formation of crystalline seeds, we adopt bond orientational analysis to detect which particles are in crystalline environments~\cite{russo2013interplay}. 
With this analysis we reveal that crystallization events indeed start occurring for the state point $\phi\approx 0.30$ and $c_p=0.38$ mg/g, which has the lowest
polymer concentration of all state points. This is shown in Fig.~\ref{fig:crystals}a, where the black line indicates the average size of the crystallites, 
while the red line indicates the average number of crystallites as a function of time. The number of crystallites first rapidly increases as nucleation events
start occurring inside the liquid network, but eventually starts decreasing as the different crystallites grow and merge with each other.
%The average size of the crystals shows an interesting linear growth regime for $100<t<600$. This linear growth regime goes past the time
%of coalescence of nuclei, so a first speculation is that it could be due to the compensation of the following two effects:
%\begin{itemize}
% \item diffusive growth of the nuclei, which scales like $t^{1/2}$
% \item coalescence growth, which should scale faster than $t$, maybe $t^2?$
%\end{itemize}
All other state points show only negligible signs of crystallization. In particular, for state points with $\phi\approx 0.30$, increasing
polymer concentration drastically reduces the amount of crystals. This is in agreement with the
idea of enhanced crystallization rates near metastable critical points~\cite{ten1997enhancement,olmsted1998spinodal} and the
results of Refs.~\cite{soga1999metastable,fortini2008crystallization,perez2011pathways},
which speculated two different arrest mechanisms: crystallization at low polymer concentration, and dynamic arrest at high polymer concentrations.

While nucleation always occurs inside the liquid branches of the phase-separating network, we suggest here the presence of a new growth channel
which was neither considered nor observed before. In the following we will show that the growth mechanism of the crystal is not only due to filling of the spinodal liquid network, as was observed in Refs.~\cite{soga1999metastable,fortini2008crystallization,perez2011pathways} for Brownian Dynamics simulations, but also involves 
evaporation of the liquid phase and subsequent direct condensation of the gas phase on the crystalline seeds. 
To confirm the presence of the two processes, we plot in Fig.~\ref{fig:crystals}b the gyration radius of individual crystalline nuclei for the state point 
$\phi\approx 0.30$ and $c_p=0.38$ mg/g. The results show that the crystal growth follows two different scaling laws: at small crystalline sizes it scales with 
a fractal dimension close to random percolation ($D=2.53$), while at large sizes it scales as $D=3$, as in compact crystal growth. This demonstrates that, while 
small crystalline nuclei are nucleated inside the phase-separated liquid branches, once they reach the transverse size of the liquid branch, the growth 
follows a volume growth with the formation of isotropic crystal droplets. The same scaling law was suggested in the analysis of the
structure factors, Fig.~\ref{fig:phase_separation}c, but here it is shown directly by the analysis of the gyration radius.

\textcolor{red}{Next we analyse in detail the crystallization trajectory, distinguishing between the different process responsible for the
crystallization of the phase separated network.
Supplementary Movie 2 shows the confocal microscopy video for the crystallizing sample at $\phi\approx 0.30$ and $c_p=0.38$ mg/g,
where the process of crystal growth by direct attachment of gas particles can be observed directly. 
From the trajectory, we extract slabs of 10 particle diameter thickness for configurations at different times, shown in Fig.~\ref{fig:transitions}a. 
Particles are coloured according to their phases (gas, liquid, and crystal). For the classification criteria see the \emph{Methods} section. The figure shows the first nucleation 
events inside the liquid network (left panel). At the same time liquid regions that have not crystallized start evaporating, and the gas phase contributes to the growth of nuclei (middle panel). Finally the different nuclei coalesce (right panel). Note that the network pattern changes between $t=2055\,\tau_B$ (left panel) and $t=5577,\tau_B$ (middle panel): not all liquid
branches of the network crystallize due to evaporation. The crystal-gel structure instead persists, as seen by comparing $t=5577,\tau_B$ (middle panel) and $t=12098,\tau_B$ (right panel).}

\textcolor{red}{
Next we show in Fig.~\ref{fig:transitions}b the fraction of particles in each different phase for the crystallization trajectory at
$\phi\approx 0.30$ and $c_p=0.38$ mg/g, after the liquid-gas phase separation has occurred.
For the classification criteria see \emph{Methods}.
The process of crystallization is characterized 
by a steep decrease in liquid particles, as they transform into small crystalline nuclei inside the liquid domains. This decrease
is then accompanied by an increase in the fraction of gas particles: as the first crystals start to reach the gas phase, liquid particles evaporate 
to the gas phase due to the higher vapour pressure of the liquid phase compared to the crystalline phase. After the onset 
of a steady-state gas population, there are three growth channels for the crystal. Direct crystallization is the process by which 
crystals grow by incorporating nearby liquid particles. In the Bergeron process, liquid droplets first evaporate and the resulting gas phase 
contributes to the growth of crystalline regions. Ostwald ripening is instead the process by which small crystallites sublimate, and colloidal
particles are transferred to larger nuclei. In order to assess the relative importance of these three growth channels, we analyse the trajectory, and, for every time frame, assign each particle a state between gas, liquid and crystal. In order to minimize short-term fluctuations, the state of each colloidal particle is time averaged for $50\,\tau_B$. We then measure the fraction of trajectories with different transition histories.
Direct crystallization accounts for 54\% of particles trajectories in which
gas or liquid particles transition to the crystal state without liquid evaporation or crystal de-sublimation. The Bergeron process accounts for 35\% of particle trajectories in which liquid particles
transition to the gas state before crystallizing. Ostwald ripening accounts for 11\% of particle trajectories in which crystal particles transition to the gas state before returning
to the crystal state. The different crystallization channels are depicted in the diagram of Fig.~\ref{fig:transitions}c.}

@@

%\subsection*{Discussion and conclusions}
The first stages of gelation always involve spinodal decomposition with the formation
of liquid network by viscoelastic gas-liquid phase separation. 
% Due to hydrodynamic interaction, the network is locally stretched, and the accumulation of mechanical stress on some bonds causes some
% of them to break and the network to reorganize in a more compact and larger structure. This occurs only at low polymer concentration,
% while at high polymer concentrations the bonds are too strong to break. This will eventually dictate which arrest mechanism
% is available to the system.
Depending on the polymer concentration, there are two possible arrest mechanisms. 
(a) Crystallization, observed at $\phi\approx 0.30$ and $c_p=0.38$ mg/g, small crystalline nuclei appear inside the liquid network, reach the surface 
 of the liquid branches, and then grow by addition of particles from the gas phase. The final structure is a network of crystal droplets,
 as confirmed by the fractal dimension of the branches, the volume fraction of the particles within the branches, and bond orientational analysis. 
(b) Dynamic arrest, particle arrest when the dynamics inside the liquid branch becomes slow, which should happen at the intersection of the
 glass line with the liquid side of the coexistence curve.

@@

We use \textsc{pmma} (poly(methyl methacrylate)) colloids sterically stabilized with methacryloxypropyl terminated \textsc{pdms}(poly(dimethyl siloxane)) and fluorescently labelled with rhodamine isothiocyanate chemically bonded to the \textsc{pmma}. 
The average colloid diameter is 2.2 $\mu$m, measured via the diffusion constant in dilute conditions without polymer. The Brownian time is $\tau_B \approx 2.3$ s. 
We assess that the size distribution of our particle is Gaussian with a polydispersity below 5\% via direct confocal measurements~\cite{Leocmach2013}.
This small polydispersity allows crystallization.
We disperse the particles in refractive index and density matching mixture of cis-decalin (Tokyo Kasei) and bromocyclohexane (Sigma-Aldrich). 
The precision of density matching is $\sim 10^{-4}$ mg/ml, for which the gravity effect on our gelation processes is negligible up to 12 hours.

To induce short-ranged depletion attraction, we use polystyrene (TOSOH) of molecular weight 3.8 Mdalton as non-adsorbing polymer. 
Experiments are conducted at 27 $^\circ$C, some 80 $^\circ$C above the theta temperature in this solvents mixture~\cite{Royall2007}. A Flory scaling of the measurements of~\cite{lu2008gelation} yields a radius of gyration $R_g=82$ nm, and thus the polymer-colloid size ratio is $q_R=2R_g/\sigma=0.075$ and the overlap mass fraction of polymer $2.1$~mg/g.

@@

To account for imprecision in particle localisation close to contact, we consider two particles bonded if their center-to-center distance is within 2.3 $\mu$m $>\sigma+2R_g$.

@@

\begin{figure*}[!t]
 \centering
 \includegraphics[width=12cm]{fig2}
 % crystal_size.eps: 0x0 pixel, 300dpi, 0.00x0.00 cm, bb=(atend)
\caption{\textcolor{red}{{\bf Percolated network structures.} {\bf a:} gel; {\bf b:} crystal-gel. The gel state is obtained from the state
point $\phi\approx 0.30$ and $c_p=1.07$ mg/g, while the crystal-gel is obtained at the same $\phi$ but
at lower polymer concentration, $c_p=0.38$ mg/g. On the left, the network is represented as a continuous field that has
been obtained by applying a gaussian filter to the position of the particles (of width equal to the inter-particle distance) and
plotting the surface which bounds the highest 20 \% values of the field. We can clearly see not only the percolated nature of both networks, 
but also the difference in the smoothness of the network surface between amorphous and crystal gels. 
On the right, slabs of 5 particle diameter thickness are plotted,
with particles coloured with the same colour scheme as in Fig.~\ref{fig:transitions}: red (crystal), green (gas particles), white (liquid particles). 
We note that very few liquid particles remain and the network is almost crystalline.}} 
\label{fig:network}
\end{figure*}

@@

\begin{figure}[!t]
 \centering
 \includegraphics{characterisation}
 % crystal_size.eps: 0x0 pixel, 300dpi, 0.00x0.00 cm, bb=(atend)
\caption{{\bf Characterization of the crystal-gel.} {\bf a,} Time evolution of the average size of the crystals (blue line) and the number of crystals (red line) for the state point $\phi\approx 0.30$ and $c_p=0.38$ mg/g. {\bf b,} Radius of gyration of crystalline nuclei at the late stages of gelation for the same state point.} 
 \label{fig:crystals}
\end{figure}

@@

\begin{figure*}[!t]
 \centering
 \includegraphics[width=14cm]{fig4}
 % snapshots.eps: 0x0 pixel, 300dpi, 0.00x0.00 cm, bb=0 -1 2256 735
 \caption{\textcolor{red}{{\bf Crystal gel formation.}  
{\bf a,} Reconstructions from confocal coordinates at $\phi\approx 0.30$ and $c_p=0.38$ mg/g. Depth of view is $5\sigma$. Particles are drawn to scale and coloured according to their phase. Red: crystal; white: liquid; green: gas. 
{\bf b,} Fraction of particles for each phase. Gray vertical lines indicate the times shown in panel a. 
{\bf c,} Each particle during its trajectory changes between three states: gas, liquid and crystal. Direct crystallization accounts for 54\% of particles trajectories in which
gas or liquid particles transition to the crystal state without liquid evaporation or crystal de-sublimation. The Bergeron process accounts for 35\% of particle trajectories in which liquid particles
transition to the gas state before crystallizing. Ostwald ripening accounts for 11\% of particle trajectories in which crystal particle transition to the gas state before returning
to the crystal state.}}
\label{fig:transitions}
\end{figure*}

>>>>

In our experimental setup, the sample cell is put in contact with a reservoir of a salt solution through a semi-permeable membrane (see Methods 
on the details of our experiments).
At $t=0$, the increase in the concentration of salt ions within a few Brownian times ($\tau_B$) screens the Coulombic repulsion between colloidal particles,
which are then subject to attractive depletion forces, making the system thermodynamically unstable and leading to phase separation. 
%In the following we discuss all the stages of the formation of the crystal gels: phase separation, stress-driven agein and crystallization.
The experimental data are taken at two different volume fractions ($\phi\approx 0.14$ and $\phi\approx 0.33$) and for different values
of the polymer concentration, $c_p=0.38,0.48,0.57,1.07$ mg/g for $\phi\approx 0.33$ and $c_p=0.82,1.36$ mg/g for $\phi\approx 0.14$.
In Fig.~\ref{fig:phase_separation}a, we superimpose the state points with a theoretical phase diagram, which we calculated from generalised free volume theory~\cite{Fleer2008} using the experimental values for the size of the colloids, the radius of gyration of the polymer, and the polymer concentration (see Supplementary Methods).
For each state point, scans were made at early times (every $10$~s, i.e. $4\tau_B$) and at the late stage (every $30$~s) of the gelation process.

@@

As described above and also in Methods, salt injection initiates liquid-gas phase separation of the colloidal suspension. 
All samples share the same early stages of spinodal decomposition. 
A typical phase ordering process observed at $\phi\approx 0.33$ and $c_p=0.38$ mg/g can be seen in Supplementary Movies 1 and 2. 
Due to strong dynamical asymmetry between colloids and the solvent \cite{tanaka1999colloid}, the colloidal particles start aggregating, 
eventually forming a percolating network. Thus, a dense network (liquid) coexists with freely diffusing monomers (gas).
In Fig.~\ref{fig:phase_separation}a, the state points which did not phase separate are indicated with open triangles, while
the other symbols indicate samples undergoing phase separation. The (black) continuous line is the gas spinodal computed from free volume theory,
and it is in good agreement with our experimental points (the small discrepancy at small colloidal volume fractions 
is common in free volume theory)~\cite{Royall2007,lu2008gelation}. 
The topology of the network is similar to all our samples at the initial stages. 
This is shown for example in Fig.~\ref{fig:phase_separation}b, where we plot the radius of gyration of colloidal clusters ($R_g$)
as a function of the cluster size ($s$) at the percolation point. Two particles belong to the same cluster if their distance is
within the first peak of the radial distribution function. All samples show the same scaling law, $R^2_{\rm g}\sim s^{2/D}$, which is compatible with
the random gelation universality class exponent ($D=2.53$) as shown by the dashed line in Fig.~\ref{fig:phase_separation}b
(we do not imply that the formation of the network follows the random gelation universality class, just that our
fractal dimension is compatible with it).

@@

While sharing similar structural properties at early times, the samples show significantly different behaviours in the later stages of the phase separation process.
To examine this in more detail, we plot in Fig.~\ref{fig:phase_separation}c the structure factor for the final stages of the gelation process for all state points.
Calculation of the structure factor is done with the Hanning window function, to minimize boundary effects.
The figure shows that at low wavenumber $q$, the structure factor displays fractal scaling compatible with Guinier law, $S(q)\sim q^{-D}$.
But a difference in the fractal dimension $D$ between the $\phi\approx 0.33$ and $c_p=0.38$ mg/g state point and other state points starts to become visible. In fact, while states
with high polymer concentration retain the exponent $D=2.5(3)$, 
which is the random gel universality class exponent, as we also observed in the early stages of the gelation process (see Fig.~\ref{fig:phase_separation}b),
the state point with low polymer concentration ($\phi\approx 0.33$ and $c_p=0.38$ mg/g) displays the biggest deviation from that exponent. In the figure we also
plot the exponent $D=3$, which corresponds to volume growth, which we will show in the following analysis to be the correct exponent for this state. 
The difference is due to the onset of crystallization in the low-polymer concentration sample. This is already evident in the high $q$ behaviour of
the structure factors (Fig.~\ref{fig:phase_separation}c), where the diffraction planes appear as sharper peaks for $\phi\approx 0.33$ and $c_p=0.38$ mg/g.

To gain additional insight, in Fig.~\ref{fig:phase_separation}d we plot the average local volume fraction as a function of the number of neighbours.
\textcolor{red}{The volume fraction is obtained by computing the Voronoi diagram of each configuration,
which uniquely assigns a volume to each colloidal particle.}
For all state points, we can see that the average volume fraction of a colloidal
particle increases with its number of neighbours. Approaching close packing (12 neighbours), we clearly see two families of curves. The state point
at $\phi\approx 0.33$ and $c_p=0.38$ mg/g reaches close packing at a volume fraction of around $\sim 68\%$, which is in reasonable agreement with the
volume fraction of the stable crystal phase~\cite{ilett1995phase}. All other state points, instead, reach close packing at a volume fraction $\sim 64\%$, which is
indeed close to the volume fraction of the attractive glass state~\cite{pham2002multiple}. 
%(\textbf{the absolute values of these volume fraction are not totally convincing, probably because of a not optimal estimate of the size of the colloids. We should run the code of Mathieau with size detection}). 
This result clearly indicates that the state point at $\phi\approx 0.33$ and $c_p=0.38$ mg/g follows a different arrest mechanism, in which phase separation
is arrested by crystallization and not by glassiness. Although our estimation of the absolute volume fraction may involves systematic errors of a few \%, the trend should be robust~\cite{Poon2012}.

@@

A direct analysis of the colloidal positions at long times confirms that indeed the new structure obtained at
$\phi\approx 0.33$ and $c_p=0.38$ mg/g has different morphological properties than usual colloidal gels.
We directly compare the two structures in Fig.~\ref{fig:network}, where Gaussian filtering is used to depict the network structure as a continuous field.
Panel a presents the results for high
polymer concentration gel, $c_p=1.07$ mg/g; panel b presents the results for the low polymer concentration sample, $c_p=0.38$ mg/g.
The comparison shows that the new arrest mechanism produces a network structure with bigger pores, smoother surfaces, and thicker strands (panel b) compared
to the colloidal gel sample (panel a). The bottom row shows the individual particle positions for slabs with a thickness of 5 particle diameters.
From the figure it is immediately clear that the network strands at low polymer concentration are crystalline. The new arrest mechanism thus
involve the formation of a crystal-gel network. In order to explain the morphological differences between colloidal gels and crystal-gels,
we now consider the kinetics of formation in our samples.

To confirm the formation of crystalline seeds, we adopt bond orientational analysis to detect which particles are in crystalline environments~\cite{russo2013interplay}. 
With this analysis we reveal that crystallization events indeed start occurring for the state point $\phi\approx 0.33$ and $c_p=0.38$ mg/g, which has the lowest
polymer concentration of all state points. This is shown in Fig.~\ref{fig:crystals}a, where the black line indicates the average size of the crystallites, 
while the red line indicates the average number of crystallites as a function of time. The number of crystallites first rapidly increases as nucleation events
start occurring inside the liquid network, but eventually starts decreasing as the different crystallites grow and merge with each other.
%The average size of the crystals shows an interesting linear growth regime for $100<t<600$. This linear growth regime goes past the time
%of coalescence of nuclei, so a first speculation is that it could be due to the compensation of the following two effects:
%\begin{itemize}
% \item diffusive growth of the nuclei, which scales like $t^{1/2}$
% \item coalescence growth, which should scale faster than $t$, maybe $t^2?$
%\end{itemize}
All other state points show only negligible signs of crystallization. In particular, for state points with $\phi\approx 0.33$, increasing
polymer concentration drastically reduces the amount of crystals. This is in agreement with the
idea of enhanced crystallization rates near metastable critical points~\cite{ten1997enhancement,olmsted1998spinodal} and the
results of Refs.~\cite{soga1999metastable,fortini2008crystallization,perez2011pathways},
which speculated two different arrest mechanisms: crystallization at low polymer concentration, and dynamic arrest at high polymer concentrations.

While nucleation always occurs inside the liquid branches of the phase-separating network, we suggest here the presence of a new growth channel
which was neither considered nor observed before. In the following we will show that the growth mechanism of the crystal is not only due to filling of the spinodal liquid network, as was observed in Refs.~\cite{soga1999metastable,fortini2008crystallization,perez2011pathways} for Brownian Dynamics simulations, but also involves 
evaporation of the liquid phase and subsequent direct condensation of the gas phase on the crystalline seeds. 
To confirm the presence of the two processes, we plot in Fig.~\ref{fig:crystals}b the gyration radius of individual crystalline nuclei for the state point 
$\phi\approx 0.33$ and $c_p=0.38$ mg/g. The results show that the crystal growth follows two different scaling laws: at small crystalline sizes it scales with 
a fractal dimension close to random percolation ($D=2.53$), while at large sizes it scales as $D=3$, as in compact crystal growth. This demonstrates that, while 
small crystalline nuclei are nucleated inside the phase-separated liquid branches, once they reach the transverse size of the liquid branch, the growth 
follows a volume growth with the formation of isotropic crystal droplets. The same scaling law was suggested in the analysis of the
structure factors, Fig.~\ref{fig:phase_separation}c, but here it is shown directly by the analysis of the gyration radius.

\textcolor{red}{Next we analyse in detail the crystallization trajectory, distinguishing between the different process responsible for the
crystallization of the phase separated network.
Supplementary Movie 2 shows the confocal microscopy video for the crystallizing sample at $\phi\approx 0.33$ and $c_p=0.38$ mg/g,
where the process of crystal growth by direct attachment of gas particles can be observed directly. 
From the trajectory, we extract slabs of 10 particle diameter thickness for configurations at different times, shown in Fig.~\ref{fig:transitions}a. 
Particles are coloured according to their phases (gas, liquid, and crystal). For the classification criteria see the \emph{Methods} section. The figure shows the first nucleation 
events inside the liquid network (left panel). At the same time liquid regions that have not crystallized start evaporating, and the gas phase contributes to the growth of nuclei (middle panel). Finally the different nuclei coalesce (right panel). Note that the network pattern changes between $t=2055\,\tau_B$ (left panel) and $t=5577,\tau_B$ (middle panel): not all liquid
branches of the network crystallize due to evaporation. The crystal-gel structure instead persists, as seen by comparing $t=5577,\tau_B$ (middle panel) and $t=12098,\tau_B$ (right panel).}

\textcolor{red}{
Next we show in Fig.~\ref{fig:transitions}b the fraction of particles in each different phase for the crystallization trajectory at
$\phi\approx 0.33$ and $c_p=0.38$ mg/g, after the liquid-gas phase separation has occurred.
For the classification criteria see \emph{Methods}.
The process of crystallization is characterized 
by a steep decrease in liquid particles, as they transform into small crystalline nuclei inside the liquid domains. This decrease
is then accompanied by an increase in the fraction of gas particles: as the first crystals start to reach the gas phase, liquid particles evaporate 
to the gas phase due to the higher vapour pressure of the liquid phase compared to the crystalline phase. After the onset 
of a steady-state gas population, there are three growth channels for the crystal. Direct crystallization is the process by which 
crystals grow by incorporating nearby liquid particles. In the Bergeron process, liquid droplets first evaporate and the resulting gas phase 
contributes to the growth of crystalline regions. Ostwald ripening is instead the process by which small crystallites sublimate, and colloidal
particles are transferred to larger nuclei. In order to assess the relative importance of these three growth channels, we analyse the trajectory, and, for every time frame, assign each particle a state between gas, liquid and crystal. In order to minimize short-term fluctuations, the state of each colloidal particle is time averaged for $50\,\tau_B$. We then measure the fraction of trajectories with different transition histories.
Direct crystallization accounts for 54\% of particles trajectories in which
gas or liquid particles transition to the crystal state without liquid evaporation or crystal de-sublimation. The Bergeron process accounts for 35\% of particle trajectories in which liquid particles
transition to the gas state before crystallizing. Ostwald ripening accounts for 11\% of particle trajectories in which crystal particles transition to the gas state before returning
to the crystal state. The different crystallization channels are depicted in the diagram of Fig.~\ref{fig:transitions}c.}

@@

%\subsection*{Discussion and conclusions}
The first stages of gelation always involve spinodal decomposition with the formation
of liquid network by viscoelastic gas-liquid phase separation. 
% Due to hydrodynamic interaction, the network is locally stretched, and the accumulation of mechanical stress on some bonds causes some
% of them to break and the network to reorganize in a more compact and larger structure. This occurs only at low polymer concentration,
% while at high polymer concentrations the bonds are too strong to break. This will eventually dictate which arrest mechanism
% is available to the system.
Depending on the polymer concentration, there are two possible arrest mechanisms. 
(a) Crystallization, observed at $\phi\approx 0.33$ and $c_p=0.38$ mg/g, small crystalline nuclei appear inside the liquid network, reach the surface 
 of the liquid branches, and then grow by addition of particles from the gas phase. The final structure is a network of crystal droplets,
 as confirmed by the fractal dimension of the branches, the volume fraction of the particles within the branches, and bond orientational analysis. 
(b) Dynamic arrest, particle arrest when the dynamics inside the liquid branch becomes slow, which should happen at the intersection of the
 glass line with the liquid side of the coexistence curve.

@@

We use \textsc{pmma} (poly(methyl methacrylate)) colloids sterically stabilized with methacryloxypropyl terminated \textsc{pdms}(poly(dimethyl siloxane)) and fluorescently labelled with rhodamine isothiocyanate chemically bonded to the \textsc{pmma}.
The diffusion constant in dilute conditions without polymer allows us to estimate the colloid diameter to $2.3\pm 0.05\, \mu$m. The Brownian time is $\tau_B \approx 2.3\,$s. 
We assess that the size distribution of our particle is Gaussian with a polydispersity below 5\% via direct confocal measurements~\cite{Leocmach2013}.
This small polydispersity allows crystallization.
We disperse the particles in refractive index and density matching mixture of cis-decalin (Tokyo Kasei) and bromocyclohexane (Sigma-Aldrich). 
The precision of density matching is $\sim 10^{-4}$ mg/ml, for which the gravity effect on our gelation processes is negligible up to 12 hours.

To induce short-ranged depletion attraction, we use polystyrene (TOSOH) of molecular weight 3.8 Mdalton as non-adsorbing polymer.
Experiments are conducted at 27 $^\circ$C, some 80 $^\circ$C above the theta temperature in this solvents mixture~\cite{Royall2007}. A Flory scaling of the measurements of~\cite{lu2008gelation} yields a radius of gyration $R_g=76\pm5$ nm.

We obtain the best fit between theoretical phase diagram and experimental data for $\sigma=2.25\,\mu$m and $R_g=80$ nm (see Supplementary method). Therefore the polymer-colloid size ratio is $q_R=2R_g/\sigma=0.071$ and the overlap mass fraction of polymer $2.25$~mg/g.

@@

To account for imprecision in particle localisation close to contact, we consider two particles bonded if their center-to-center distance is within 2.3 $\mu$m $\approx\sigma+2R_g$.

@@

\begin{figure*}[!t]
 \centering
 \includegraphics[width=12cm]{fig2}
 % crystal_size.eps: 0x0 pixel, 300dpi, 0.00x0.00 cm, bb=(atend)
\caption{\textcolor{red}{{\bf Percolated network structures.} {\bf a:} gel; {\bf b:} crystal-gel. The gel state is obtained from the state
point $\phi\approx 0.33$ and $c_p=1.07$ mg/g, while the crystal-gel is obtained at the same $\phi$ but
at lower polymer concentration, $c_p=0.38$ mg/g. On the left, the network is represented as a continuous field that has
been obtained by applying a gaussian filter to the position of the particles (of width equal to the inter-particle distance) and
plotting the surface which bounds the highest 20 \% values of the field. We can clearly see not only the percolated nature of both networks, 
but also the difference in the smoothness of the network surface between amorphous and crystal gels. 
On the right, slabs of 5 particle diameter thickness are plotted,
with particles coloured with the same colour scheme as in Fig.~\ref{fig:transitions}: red (crystal), green (gas particles), white (liquid particles). 
We note that very few liquid particles remain and the network is almost crystalline.}} 
\label{fig:network}
\end{figure*}

@@

\begin{figure}[!t]
 \centering
 \includegraphics{characterisation}
 % crystal_size.eps: 0x0 pixel, 300dpi, 0.00x0.00 cm, bb=(atend)
\caption{{\bf Characterization of the crystal-gel.} {\bf a,} Time evolution of the average size of the crystals (blue line) and the number of crystals (red line) for the state point $\phi\approx 0.33$ and $c_p=0.38$ mg/g. {\bf b,} Radius of gyration of crystalline nuclei at the late stages of gelation for the same state point.} 
 \label{fig:crystals}
\end{figure}

@@

\begin{figure*}[!t]
 \centering
 \includegraphics[width=14cm]{fig4}
 % snapshots.eps: 0x0 pixel, 300dpi, 0.00x0.00 cm, bb=0 -1 2256 735
 \caption{\textcolor{red}{{\bf Crystal gel formation.}  
{\bf a,} Reconstructions from confocal coordinates at $\phi\approx 0.33$ and $c_p=0.38$ mg/g. Depth of view is $5\sigma$. Particles are drawn to scale and coloured according to their phase. Red: crystal; white: liquid; green: gas. 
{\bf b,} Fraction of particles for each phase. Gray vertical lines indicate the times shown in panel a. 
{\bf c,} Each particle during its trajectory changes between three states: gas, liquid and crystal. Direct crystallization accounts for 54\% of particles trajectories in which
gas or liquid particles transition to the crystal state without liquid evaporation or crystal de-sublimation. The Bergeron process accounts for 35\% of particle trajectories in which liquid particles
transition to the gas state before crystallizing. Ostwald ripening accounts for 11\% of particle trajectories in which crystal particle transition to the gas state before returning
to the crystal state.}}
\label{fig:transitions}
\end{figure*}
